---
title: "Ionflow: Ionomics data network and enrichment analysis"
author: "Wanchang Lin"
date: 01-12-2020
KeyWords: ionomics, network analysis, enrichment analysis 
description: |
  Ionomics data gene network and enrichment analysis
fontsize: 11pt
geometry: margin=1in
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 7, 
                      fig.align = "center",
                      fig.path = "./img/",
                      dpi = 100, dev = "png", cache = FALSE)

pkgs <- c("reshape2", "plyr", "knitr", "kableExtra", "tidyverse",
          "ggrepel", "corrplot", "gplots", "network", "sna", "GGally",
          "org.Sc.sgd.db", "GO.db", "GOstats", "KEGG.db", "pheatmap")
invisible(lapply(pkgs, library, character.only = TRUE))
source("ionflow_funcs.R")
```

This document explains how to performs ionomics data analysis including gene
network and enrichment analysis. 

## Data preparation

To explore the pipeline, we'll use the ionomics data set:

```{r}
ion_data <- read.table("./test-data/iondata.tsv", header = T, sep = "\t")
dim(ion_data)
```

Ten random lines are shown as:

```{r, eval = F, echo = T}
sample_n(ion_data, 10)
```

```{r, eval = T, echo = F}
sample_n(ion_data, 10) %>%
  kable(caption = 'Samples of raw data', digits = 2) %>% 
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

We can see that the first few columns are meta information such as gene ORF
and batch id. The rest is the ionomics data. 


## Data pre-process

The raw data set is needed to be pre-processed. This involves:

- log transformation
- batch correction
- outlier detection
- standardisation

For batch correction, control line could be used. If so, the values belong
to control lines are used to be the basis of batch correlation. This data
has a control line: **YDL227C** mutant. The code segment below is to
identify it:

```{r}
max(with(ion_data, table(Knockout)))
which.max(with(ion_data, table(Knockout)))
```

Standarisation provides a *custom* method. This allows user to use
specific std values such as: 

```{r}
std <- read.table("./test-data/user_std.tsv", header = T, sep = "\t")
std
```

The outlier detection here is univarite method, with a threshold to control
the number of outliers. The larger the threshold (`thres_outl`) the more
outlier removal. 

The pre-process procedure returns not only processed ionomics data but also
a symbolic data. This data is based on the inomics data and a
threshold(`thres_symb`):

- `0` if ionomics data located between `[-thres_symb, thres_symb]`
- `1` if ionomics data larger than `thres_symb`
- `-1` if ionomics data smaller than `-thres_symb`

Ths symbolic data is important since the network analysis will use it along
with ionomics data, and enrichment analysis will be performd only based on
it. It also should be noted that the symblic data is sensitive to the
choices of the threshold. 

Let's run the pre-process procedure:

```{r, tidy = FALSE}
pre <- PreProcessing(data = ion_data,
                     var_id = 1, batch_id = 2, data_id = 3,
                     method_norm = "median",
                     control_lines = "YDL227C",
                     control_use = "control",
                     method_outliers = "IQR",
                     thres_outl = 3,
                     stand_method = "std",
                     stdev = NULL,
                     thres_symb = 3)

names(pre)
```

The pre-processed data are returned with some summaries, one of them is the
z-score.

```{r, results = 'asis'}
pre$stats.batch_data %>% kable(caption = 'Summary: z-scores', digits = 2) %>%
  kable_styling(full_width = F, font_size = 10)
```

The pre-processed data and its symbolic data are like like:

```{r}
pre$data.gene.zscores %>% head() %>%
  kable(caption = 'Pre-processed data', digits = 2) %>% 
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))

pre$data.gene.symb %>% head() %>%
  kable(caption = 'Symbolic data') %>%
  kable_styling(full_width = F, font_size = 10)
```

The pre-processed data distribution is:

```{r pre-proc-ion, fig.cap = "Ionome data distribution plot"}
pre$plot.hist
```

## Data filtering

There are a lot of ways to filter gene. Here we filter gene based on
symbolic data:

```{r}
data <- pre$data.gene.zscores
data_symb <- pre$data.gene.symb
idx <- rowSums(abs(data_symb[, -1])) > 0
dat <- data[idx, ]
dat_symb <- data_symb[idx, ]
dim(dat)
```

## Gene network

The gene network is based on the ionomics and symboloc data and uses the
similarity measure results to build up the network. The similarity measure
method is one of *pearson*, *spearman*, *kendall*, *cosine*, *mahal_cosine*
or *hybrid_mahal_cosine*.

First, the Pearson correlation is used to build up the network:

```{r net, fig.cap = "Netwok analysis based on Pearson correlation"}
net <- GeneNetwork(data = dat,
                   data_symb = dat_symb,
                   min_clust_size = 10,
                   thres_corr = 0.75,
                   method_corr = "pearson")
net$plot.pnet1
net$plot.pnet2
net$plot.impact_betweenness
```

The node colours are indicated by either the similarity measures or the
network community detection, i.e. clustering.

\clearpage

For the comparision, we use different similarity methods. Here use *Cosine*:

```{r net-1, fig.cap = "Netwok analysis based on Cosine"}
net_1 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "cosine")
net_1$plot.pnet1
net_1$plot.pnet2
```

\clearpage

Use *Hybrid Mahalanobis Cosine*:

```{r net-2, fig.cap = "Netwok analysis based on Mahalanobis Cosine"}
net_2 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "mahal_cosine")
net_2$plot.pnet1
net_2$plot.pnet2
```

\clearpage

Again, we use *Hybrid Mahalanobis Cosine*:

```{r net-3, fig.cap = "Netwok analysis based on Hybrid Mahalanobis Cosine"}
net_3 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "hybrid_mahal_cosine")
net_3$plot.pnet1
net_3$plot.pnet2
```

\clearpage

## Enrichment analysis

The KEGG enrichment analysis:

```{r, message = F}
kegg <- kegg_enrich(data = dat_symb, min_clust_size = 10, pval = 0.05,
                    annot_pkg =  "org.Sc.sgd.db")

#' kegg
kegg %>% 
  kable(caption = 'KEGG enrichmenat analysis', digits = 3) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

Note that there can be none results for KRGG  enrichment analysis. Change
arguments such as `thres_clus` as appropriate.

The GO Terms enrichment analysis:

```{r, message = F}
go <- go_enrich(data = dat_symb, min_clust_size = 10, pval = 0.05,
                ont = "BP", annot_pkg =  "org.Sc.sgd.db")
#' go
go %>% head() %>%
  kable(caption = 'GO Terms enrichmenat analysis', digits = 3) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

\newpage

## Exploratory analysis

Some analysis are performed in terms of ions, i.e. feature, including PCA
and correlation.

```{r expl, fig.cap = "Exploratory analysis plots with respect to ionome"}
expl <- ExploratoryAnalysis(data = dat)
expl$plot.PCA_Individual
expl$plot.correlation_network
```

