---
title: "Ionflow: network and enrichment analysis for ionomics data"
author: "Wanchang Lin"
date: 01-12-2020
KeyWords: ionomics, network analysis, enrichment analysis 
description: |
  Ionomics data gene network and enrichment analysis
output:
  BiocStyle::html_document:
    toc_depth: 3
    number_section: false
    toc_float: false
  BiocStyle::pdf_document:
    toc_depth: 3
    number_section: false
    keep_tex: false
geometry: margin=1in
fontsize: 11pt
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 7, 
                      fig.align = "center",
                      fig.path = "./img/",
                      dpi = 100, dev = "png", cache = T)

pkgs <- c("reshape2", "plyr", "knitr", "kableExtra", "tidyverse",
          "ggrepel", "corrplot", "gplots", "network", "sna", "GGally",
          "org.Sc.sgd.db", "GO.db", "GOstats", "KEGG.db", "pheatmap")
invisible(lapply(pkgs, library, character.only = TRUE))
source("./ionflow_funcs.R")
```

This vignette explains how to perform ionomics data analysis including gene
network and enrichment analysis by using a modification of the R package,
[ionflow](https://github.com/AlinaPeluso/MetaboFlow). The
modification(`ionflow_funcs`) was made by Wanchang Lin
(w.lin@imperial.ac.uk) and Jacopo Iacovacci(j.iacovacci@imperial.ac.uk).

## Data preparation

To explore the process, we'll use the ionomics data set:

```{r}
ion_data <- read.table("./test-data/iondata.tsv", header = T, sep = "\t")
dim(ion_data)
```

Ten random data records are shown as:

```{r, eval = F, echo = T}
sample_n(ion_data, 10)
```

```{r, eval = T, echo = F}
sample_n(ion_data, 10) %>%
  kable(caption = 'Samples of raw data', digits = 2, booktabs = T) %>% 
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

The first few columns are meta information such as gene ORF and batch id.
The rest is the ionomics data. 

## Data pre-processing

The raw data set should be pre-processed. The pre-processing function
`PreProcessing` has functions:

- log transformation
- batch correction
- outlier detection
- standardisation

The raw data are at first log transformed and then followed by the batch
correction. User can chose not to perform batch correction, otherwise
default will be either *median* or *median* plus *std* method. If there is
quality control for the batch correction, the user can use it and indicates
in the argument of `control_lines`. Also one argument gives the user the
option on how to use these control lines (`control_use`): If `control_use`
is `control`, these control lines (data rows) are used for the batch
correction factor; if `control.out`, others lines are used.

This data set has a control line: **YDL227C** mutant. The code segment below
is to identify it:

```{r}
max(with(ion_data, table(Knockout)))
which.max(with(ion_data, table(Knockout)))
```

The next stage is outlier detection. Here only univariate methods are
implemented, including *mad*, *IQR*, and *log.FC.dist*. And like batch
correction, the user can skip this procedure by setting `method_outliers =
none` in the function argument. There is a threshold to control the number
of outliers. The larger the threshold (`thres_outl`) the more outlier
removal. 

Standardisation provides three methods: *std*, *mad* or *custom*. If the
method is *custom*, the user uses a specific *std* file like: 

```{r}
std <- read.table("./test-data/user_std.tsv", header = T, sep = "\t")
std
```

The pre-processing procedure returns not only processed ionomics data but
also a symbolic data set. This data set is based on the ionomics data and is
determined by a threshold(`thres_symb`):

- `0` if ionomics value is located in `[-thres_symb, thres_symb]`
- `1` if ionomics value is larger than `thres_symb`
- `-1` if ionomics value is smaller than `-thres_symb`

Note that the symbolic data set is important since the key part of the
network and enrichment analysis is based on the hierarchical clustering of
symbolic data. 

Let's run the pre-process procedure:

```{r, tidy = FALSE}
pre <- PreProcessing(data = ion_data,
                     var_id = 1, batch_id = 2, data_id = 3,
                     method_norm = "median",
                     control_lines = "YDL227C",
                     control_use = "control",
                     method_outliers = "IQR",
                     thres_outl = 3,
                     stand_method = "std",
                     stdev = NULL,
                     thres_symb = 3)

names(pre)
```

The results include summaries of raw data and processed data. The latter
is:

```{r, results = 'asis'}
pre$stats.batch_data %>% 
  kable(caption = 'Processed data summary', digits = 2, booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10)
```

The pre-processed data and symbolic data are like this:

```{r}
pre$data.gene.zscores %>% head() %>%
  kable(caption = 'Processed data', digits = 2, booktabs = T) %>% 
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))

pre$data.gene.symb %>% head() %>%
  kable(caption = 'Symbolic data', booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10)
```

The symbolic data are calculated from the processed data with control of
`thres_symb` (here it is 3). You can obtain a new symbol data set by
re-assigning a new threshold to the function `symbol_data`:

```{r}
data_symb <- symbol_data(pre$data.gene.zscores, thres_symb = 2)
data_symb %>% head() %>%
  kable(caption = 'Symbolic data with threshold of 2', booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10)
```

The `thres_symb` is a crucial value to get the symbolic data. Before
re-setting this threshold, the user should check the summary of processed
data and pay attention to the maximum values. For example, some ions
(for example, *Cd* and *Mn*) are all zero even with 2 of `thres_symb`.

The pre-processed data distribution is:

```{r pre-proc-ion, fig.cap = "Ionomcs data distribution plot"}
pre$plot.hist
```

## Data filtering

There are a lot of ways to filter genes. Here genes are filtered according
to symbolic data: remove genes with all values which are zero.

```{r}
data <- pre$data.gene.zscores
data_symb <- pre$data.gene.symb
idx <- rowSums(abs(data_symb[, -1])) > 0
dat <- data[idx, ]
dat_symb <- data_symb[idx, ]
dim(dat)
```

## Data clustering

The hierarchical cluster analysis is the key part of gene network and gene
enrichment analysis. The methodology is as follow:

- Compute the distance of symbolic data
- Hierarchical cluster analysis on the distance
- Identify clusters/groups with a threshold of minimal number of cluster
  size

One example is:

```{r}
clust <- gene_clus(dat_symb[, -1], min_clust_size = 10)
names(clust)
```

The cluster centres are:

```{r}
clust$tab_sub
```

This shows clusters and the number of genes (larger than `min_cluster_size`).

The identified gene located in those clusters are:

```{r}
sum(clust$idx)                          #' numbers of all genes
head(as.character(dat[,1][clust$idx]))  #' and some are
```

## Gene network

The gene network uses both the ionomics and symbolic data. The similarity
measures on ionomics data are used to construct the network. Before creating
a network, these analyses are further filtered by:

- clustering of symbolic data;
- and the similarity threshold located between 0 and 1;

The methods implemented are: *pearson*, *spearman*, *kendall*, *cosine*,
*mahal_cosine* or *hybrid_mahal_cosine*. The first three methods are
correlation methods and *cosine* is similar to the Pearson correlation
which is the [cosine similarity between two centred
vectors](https://bit.ly/3oED6hj). For the last two methods, see publication:
[Extraction and Integration of Genetic Networks from Short-Profile Omic Data
Sets](https://www.mdpi.com/2218-1989/10/11/435/htm) for details.

For example, we use the Pearson correlation as similarity measure for
network analysis:

```{r}
net <- GeneNetwork(data = dat,
                   data_symb = dat_symb,
                   min_clust_size = 10,
                   thres_corr = 0.75,
                   method_corr = "pearson")
```

The network with nodes coloured by the symbolic data clustering is:

```{r net-symb, fig.cap = "Network with Pearson correlation: symbolic clustering"}
net$plot.pnet1
```

The same network, but nodes are coloured by the network community detection:

```{r net-clus, fig.cap = "Network with Pearson correlation: community detction"}
net$plot.pnet2
```

The network analysis also returns a network impact and betweenness plot:

```{r net-imp, fig.cap = "Network with Pearson correlation: impact and betweeness"}
net$plot.impact_betweenness
```

\clearpage

For comparison purposes, we use different similarity methods. Here we
choose *Cosine*:

```{r net-1, fig.cap = "Network analysis based on Cosine"}
net_1 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "cosine")
net_1$plot.pnet1
net_1$plot.pnet2
```

\clearpage

Use *Hybrid Mahalanobis Cosine*:

```{r net-2, fig.cap = "Network with Mahalanobis Cosine"}
net_2 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "mahal_cosine")
net_2$plot.pnet1
net_2$plot.pnet2
```

\clearpage

Again, we use *Hybrid Mahalanobis Cosine*:

```{r net-3, fig.cap = "Network with Hybrid Mahalanobis Cosine"}
net_3 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "hybrid_mahal_cosine")
net_3$plot.pnet1
net_3$plot.pnet2
```

\clearpage

## Enrichment analysis

The enrichment analysis is based on symbolic data clustering. The genes in
clusters are considered target gene sets while genes in the whole data set
is the universal gene set. The Bioconductor R package
[GOstats](http://bioconductor.org/packages/release//bioc/html/GOstats.html)
is used for the enrichment analysis.

The KEGG enrichment analysis, using a p-values of 0.05 and Genome wide
annotation for Yeast,
[org.Sc.sgd.db](https://bioconductor.org/packages/release/data/annotation/html/org.Sc.sgd.db.html):

```{r, message = F}
kegg <- kegg_enrich(data = dat_symb, min_clust_size = 10, pval = 0.05,
                    annot_pkg =  "org.Sc.sgd.db")

#' kegg
kegg %>% 
  kable(caption = 'KEGG enrichmenat analysis',
        digits = 3, booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

Note that there could be no results returned for KEGG enrichment analysis.
Arguments such as `min_clust_size` can be changed as appropriate.

The GO Terms enrichment analysis with ontology of *BP* (other two are *MF*
and *CC*):

```{r, message = F}
go <- go_enrich(data = dat_symb, min_clust_size = 10, pval = 0.05,
                ont = "BP", annot_pkg =  "org.Sc.sgd.db")
#' go
dim(go)
go %>% head() %>% 
  kable(caption = 'GO Terms enrichmenat analysis',
        digits = 3, booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

\clearpage

## Exploratory analysis

The exploratory analysis performs PCA and correlation analysis for ions in
terms of genes. Note that this analysis treats ions as samples/replicates
while genes are treated as variables/features. The exploratory analysis is
initially employed at an early stage of the analysis.

For example, we apply it to the pre-processed data `dat` before any other
analysis:

```{r expl-1, fig.show = "hide"}
expl <- ExploratoryAnalysis(data = dat)
names(expl)
```

The PCA plot is:

```{r dat-pca, fig.cap = "Ion PCA plot on pre-processed data"}
expl$plot.pca
```

The Person correlation of ions are shown in correlation plot, heatmap and
network plot:

```{r dat-corr, fig.cap = "Ion correlation plots on pre-processed data"}
expl$plot.corr
expl$plot.corr.heat
expl$plot.net
```

The correlation between ions and genes are shown in heatmap with dendrogram:

```{r dat-corr-gene, fig.cap = "Correlation between ions and genes on pre-processed data"}
expl$plot.heat
```

\clearpage

The exploratory analysis can also be used at other stages of the analysis.
Here for example after gene clustering analysis:

```{r expl-2, fig.cap = "Exploratory analysis after gene clustering"}
#' update data set with results of gene clustering
dat_clus <- dat[clust$idx, ]
dim(dat_clus)

expl.1 <- ExploratoryAnalysis(data = dat_clus)
expl.1$plot.pca
expl.1$plot.net
```
