---
title: "Ionflow: Ionomics data network and enrichment analysis"
author: "Wanchang Lin"
date: 01-12-2020
KeyWords: ionomics, network analysis, enrichment analysis 
description: |
  Ionomics data gene network and enrichment analysis
fontsize: 11pt
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 7, 
                      fig.align = "center",
                      fig.path = "./img/",
                      dpi = 100, dev = "png", cache = FALSE)

pkgs <- c("reshape2", "plyr", "knitr", "kableExtra", "tidyverse",
          "ggrepel", "corrplot", "gplots", "network", "sna", "GGally",
          "org.Sc.sgd.db", "GO.db", "GOstats", "KEGG.db", "pheatmap")
invisible(lapply(pkgs, library, character.only = TRUE))
source("../ionflow_funcs.R")
```

This document explains how to performs ionomics data analysis including gene
network and enrichment analysis. 

## Data preparation

To explore the pipeline, we'll use the ionomics data set:

```{r}
ion_data <- read.table("../test-data/iondata.tsv", header = T, sep = "\t")
dim(ion_data)
```

Ten random lines are shown as:

```{r, eval = F, echo = T}
sample_n(ion_data, 10)
```

```{r, eval = T, echo = F}
sample_n(ion_data, 10) %>%
  kable(caption = 'Samples of raw data', digits = 2, booktabs = T) %>% 
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

We can see that the first few columns are meta information such as gene ORF
and batch id. The rest is the ionomics data. 


## Data pre-process

The raw data set is needed to be pre-processed. The pre-processing function
`PreProcessing` performs:

- log transformation
- batch correction
- outlier detection
- standardisation

For batch correction, control line could be used. If so, the values belong
to control lines are used to be the basis of batch correlation. This data
has a control line: **YDL227C** mutant. The code segment below is to
identify it:

```{r}
max(with(ion_data, table(Knockout)))
which.max(with(ion_data, table(Knockout)))
```

The outlier detection here is univarite method, with a threshold to control
the number of outliers. The larger the threshold (`thres_outl`) the more
outlier removal. 

Standarisation provides a *custom* method. This allows user to use
specific std values such as: 

```{r}
std <- read.table("../test-data/user_std.tsv", header = T, sep = "\t")
std
```

The pre-process procedure returns not only processed ionomics data but also
a symbolic data. This data is based on the inomics data and a
threshold(`thres_symb`):

- `0` if ionomics data located between `[-thres_symb, thres_symb]`
- `1` if ionomics data larger than `thres_symb`
- `-1` if ionomics data smaller than `-thres_symb`

The core part of network and enrivhment analysis, clustering, is based on
the symbolic data. Note that the symblic data is sensitive to the choices of
the threshold. 

Let's run the pre-process procedure:

```{r, tidy = FALSE}
pre <- PreProcessing(data = ion_data,
                     var_id = 1, batch_id = 2, data_id = 3,
                     method_norm = "median",
                     control_lines = "YDL227C",
                     control_use = "control",
                     method_outliers = "IQR",
                     thres_outl = 3,
                     stand_method = "std",
                     stdev = NULL,
                     thres_symb = 3)

names(pre)
```

The results includes summaries of raw data and processed data. The latter
is:

```{r, results = 'asis'}
pre$stats.batch_data %>% 
  kable(caption = 'Processed data summary', digits = 2, booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10)
```

The pre-processed data and its symbolic data are like like:

```{r}
pre$data.gene.zscores %>% head() %>%
  kable(caption = 'Pre-processed data', digits = 2, booktabs = T) %>% 
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))

pre$data.gene.symb %>% head() %>%
  kable(caption = 'Symbolic data', booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10)
```

The symbolic data is calulated from processed data with control of
`thres_symb` (here is 3). You can obtain a new symbol data by assigning
new threshold to the function `symbol_data`:

```{r}
data_symb <- symbol_data(pre$data.gene.zscores, thres_symb = 2)
data_symb %>% head() %>%
  kable(caption = 'Symbolic data with threshold of 2', booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10)
```

The pre-processed data distribution is:

```{r pre-proc-ion, fig.cap = "Ionome data distribution plot"}
pre$plot.hist
```

## Data filtering

There are a lot of ways to filter gene. Here we filter gene based on
symbolic data:

```{r}
data <- pre$data.gene.zscores
data_symb <- pre$data.gene.symb
idx <- rowSums(abs(data_symb[, -1])) > 0
dat <- data[idx, ]
dat_symb <- data_symb[idx, ]
dim(dat)
```

## Data clustering

The hierarchical cluster analysis is the key part of gene network and gene
enrichment analysis. The methodology is as follow:

- Compute the distance of symbolic data
- Hierarchical cluster analysis on the distance
- Identify clusters/groups with a threshold of minimal number of cluster
  size

One example is:

```{r}
clust <- gene_clus(dat_symb[, -1], min_clust_size = 10)
names(clust)
```

The cluster centres are:

```{r}
clust$tab_sub
```

It indicates that clusters and their number of genes (larger than
`min_cluster_size`).

## Gene network

The gene network uses both the ionomics and symboloc data. The similarity
measure on the ionomics data is filtered by the similarity threshold located
between 0 and 1, and cluster centres of symbolic data. The filter values are
then used for network analysis. 

The similarity measure method is one of *pearson*, *spearman*, *kendall*,
*cosine*, *mahal_cosine* or *hybrid_mahal_cosine*.

First, the Pearson correlation is used to build up the network:

```{r}
net <- GeneNetwork(data = dat,
                   data_symb = dat_symb,
                   min_clust_size = 10,
                   thres_corr = 0.75,
                   method_corr = "pearson")
```

The network with nodes colouring by the symbolic clustering is:

```{r net-symb, fig.cap = "Netwok analysis based on Pearson correlation: symbolic clustering"}
net$plot.pnet1
```

The same network, but nodes are colured by the netwok community detection:

```{r net-clus, fig.cap = "Netwok analysis based on Pearson correlation: community detction"}
net$plot.pnet2
```

The network analysis also returns a network impact and betweeness plot:

```{r net-imp, fig.cap = "Netwok analysis based on Pearson correlation: impact and betweeness"}
net$plot.impact_betweenness
```

\clearpage

For the comparision purpose, we use different similarity methods. Here use
*Cosine*:

```{r net-1, fig.cap = "Netwok analysis based on Cosine"}
net_1 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "cosine")
net_1$plot.pnet1
net_1$plot.pnet2
```

\clearpage

Use *Hybrid Mahalanobis Cosine*:

```{r net-2, fig.cap = "Netwok analysis based on Mahalanobis Cosine"}
net_2 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "mahal_cosine")
net_2$plot.pnet1
net_2$plot.pnet2
```

\clearpage

Again, we use *Hybrid Mahalanobis Cosine*:

```{r net-3, fig.cap = "Netwok analysis based on Hybrid Mahalanobis Cosine"}
net_3 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = 10,
                     thres_corr = 0.75,
                     method_corr = "hybrid_mahal_cosine")
net_3$plot.pnet1
net_3$plot.pnet2
```

\clearpage

## Enrichment analysis

The KEGG enrichment analysis:

```{r, message = F}
kegg <- kegg_enrich(data = dat_symb, min_clust_size = 10, pval = 0.05,
                    annot_pkg =  "org.Sc.sgd.db")

#' kegg
kegg %>% 
  kable(caption = 'KEGG enrichmenat analysis', digits = 3, booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

Note that there can be none results for KRGG  enrichment analysis. Change
arguments such as `thres_clus` as appropriate.

The GO Terms enrichment analysis:

```{r, message = F}
go <- go_enrich(data = dat_symb, min_clust_size = 10, pval = 0.05,
                ont = "BP", annot_pkg =  "org.Sc.sgd.db")
#' go
go %>% head() %>%
  kable(caption = 'GO Terms enrichmenat analysis', digits = 3, booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

\newpage

## Exploratory analysis

Some analysis are performed in terms of ions, i.e. feature, including PCA
and correlation.

```{r expl, fig.cap = "Exploratory analysis plots with respect to ionome"}
expl <- ExploratoryAnalysis(data = dat)
expl$plot.PCA_Individual
expl$plot.correlation_network
```

