---
title: "Ionomics analysis for human data set using Ionflow"
author: "Wanchang Lin"
date: "11-12-2020, Fri"
output:
  BiocStyle::html_document:
    toc_depth: 3
    number_section: false
    toc_float: false
  BiocStyle::pdf_document:
    toc_depth: 3
    number_section: false
    keep_tex: false
geometry: margin=1in
fontsize: 11pt
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 7, 
                      fig.align = "center",
                      fig.path = "./img/",
                      dpi = 100, dev = "png", cache = F)

pkgs <- c("reshape2", "plyr", "knitr", "kableExtra", "tidyverse",
          "ggrepel", "corrplot", "gplots", "network", "sna", "GGally",
          "org.Hs.eg.db", "GO.db", "GOstats", "KEGG.db", "pheatmap")
invisible(lapply(pkgs, library, character.only = TRUE))
source("./ionflow_funcs.R")
```

## Data preparation

The human ionomics data set has been pre-processed. We need to get the
symbolic data:

```{r}
dat <- read.table("./test-data/human.csv", header = T, sep = ",")
dat <- dat[!duplicated(dat[, 1]), ]
colnames(dat)[1] <- "Line"
dat_symb <- symbol_data(x = dat, thres_symb = 3)
```

Some of ionomics data and symbolic data are like:

```{r}
dat %>% head(10) %>%
  kable(caption = 'Ionomics data', digits = 2, booktabs = T) %>% 
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))

dat_symb %>% head(10) %>%
  kable(caption = 'Symbolic data', booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

These data are filtered, i.e. remove all zero genes in symbolic data set:

```{r}
idx <- rowSums(abs(dat_symb[, -1])) > 0
dat <- dat[idx, ]
dat_symb <- dat_symb[idx, ]
dim(dat)
```

## Data clustering

The hierarchical cluster analysis is the key part of gene network and gene
enrichment analysis. The methodology is as follow:

- Compute the distance of symbolic data
- Hierarchical cluster analysis on the distance
- Identify clusters/groups with a threshold of minimal number of cluster
  size

One example is:

```{r}
min <- 8 
clust <- gene_clus(dat_symb[, -1], min_clust_size = min)
names(clust)
clust$tab_sub
```

## Gene network

The gene network uses both the ionomics and symbolic data. The similarity
measures on ionomics data are used to construct the network. Before creating
a network, these analyses are further filtered by:

- clustering of symbolic data;
- and the similarity threshold located between 0 and 1;

The methods implemented are: *pearson*, *spearman*, *kendall*, *cosine*,
*mahal_cosine* or *hybrid_mahal_cosine*.

We use the Pearson correlation as similarity measure for network analysis:

```{r}
net <- GeneNetwork(data = dat,
                   data_symb = dat_symb,
                   min_clust_size = min,
                   thres_corr = 0.6,
                   method_corr = "pearson")
```

The network with nodes coloured by the symbolic data clustering is:

```{r net-symb, fig.cap = "Network with Pearson correlation: symbolic clustering"}
net$plot.pnet1
```

The same network, but nodes are coloured by the network community detection:

```{r net-clus, fig.cap = "Network with Pearson correlation: community detction"}
net$plot.pnet2
```

The network analysis also returns a network impact and betweenness plot:

```{r net-imp, fig.cap = "Network with Pearson correlation: impact and betweenness"}
net$plot.impact_betweenness
```

\clearpage

For comparison purposes, we use *Mahalanobis Cosine*:

```{r net-2, fig.cap = "Network with Mahalanobis Cosine"}
net_2 <- GeneNetwork(data = dat,
                     data_symb = dat_symb,
                     min_clust_size = min,
                     thres_corr = 0.6,
                     method_corr = "mahal_cosine")
net_2$plot.pnet1
net_2$plot.pnet2
```

\clearpage


## Enrichment analysis

The enrichment analysis is for group data. The genes in the group are
considered target gene sets while genes in the whole data set is the
universal gene set. 

The netowk analysis retunes a vertex attributes matrix:

```{r}
head(net$net_node)
```

The second and third columns are symbolic clustering and network community
cluster, respectively. If we perform enrichment analysis on the network
community centre, the matrix should include the first column (gene IDs) and
the third colum:

```{r, message = F}
mat <- net$net_node[, c(1,3)]
kegg <- kegg_enrich(mat = mat, pval = 0.05, annot_pkg =  "org.Hs.eg.db")

#' kegg
kegg %>% 
  kable(caption = 'KEGG enrichment analysis',
        digits = 3, booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

Note that there could be no results returned for KEGG enrichment analysis.
Arguments such as `min_clust_size` can be changed as appropriate.

The GO Terms enrichment analysis with ontology of *BP* (other two are *MF*
and *CC*):

```{r, message = F}
go <- go_enrich(mat = mat, pval = 0.05, ont = "BP", annot_pkg =  "org.Hs.eg.db")
#' go
dim(go)
go %>%
  kable(caption = 'GO Terms enrichment analysis',
        digits = 3, booktabs = T) %>%
  kable_styling(full_width = F, font_size = 10,
                latex_options = c("striped", "scale_down"))
```

\clearpage

## Exploratory analysis

The explanatory analysis performs PCA and correlation analysis for ions in
terms of genes. Note that this analysis treats ions as samples/replicates
while genes are treated as variables/features. The explanatory analysis is
initially employed at an early stage of the analysis.

We apply it to the pre-processed data `dat` before any other analysis:

```{r expl-1, fig.show = "hide"}
expl <- ExploratoryAnalysis(data = dat)
names(expl)
```

The PCA plot is:

```{r dat-pca, fig.cap = "Ion PCA plot on pre-processed data"}
expl$plot.pca
```

The Person correlation of ions are shown in correlation plot, heatmap and
network plot:

```{r dat-corr, fig.cap = "Ion correlation plots on pre-processed data"}
expl$plot.corr
expl$plot.corr.heat
expl$plot.net
```

The correlation between ions and genes are shown in heatmap with dendrogram:

```{r dat-corr-gene, fig.cap = "Correlation between ions and genes on pre-processed data"}
expl$plot.heat
```
